---
title: "Books"
output: pdf_document
author: Shannon Rush
---

The goal of this analysis is to explore book genre popularity. Some questions I'll try to answer are: Have genre tastes changed over time? Does genre popularity have a pattern in the course of a year? Do demographics such as age or gender tend to report reading the same types of books?

# The Data

This data was collected via a public API from a popular book social network. All information is self reported, including demographic information, books read, and genre categorization. Unfortunately it is not necessary when reporting having read a book to categorize it by genre but fortunately many users do.

Two types of data were collected: User data, and information about books users self-reported to be "currently reading" 

A note about github: The collected CSV files are too large to share on github. However, you can see exactly how I collected the data at my repo called ['goodreads-analyses'](https://github.com/shannonrush/goodreads-analyses) I'm also happy to share the data I collected if you would like to reproduce my analyses or work with the data yourself. A word of caution: I collected this data too quickly for their tastes and ended up getting IP banned for about a month. Consume at your own risk. 

## User Data

```{r user_data, cache=TRUE}
    user.cols <- c("userID", "name", "gender", "age", "location", "lastactive", 
                   "readcount", "currentcount", "wantscount")
    users <- read.csv("data/users.csv", header=F, col.names=user.cols, stringsAsFactors=F)
    head(users)
    dim(users)
```

## Book Data

```{r book_data, cache=TRUE}
    book.cols <- c("userID", "bookID", "shelf", "dateadded", "datestarted", "datefinished", "title",
                   "isbn", "isbn13", "imageurl", "pages", "publisher", "publicationdate", "genres")
    books <- read.csv("data/currently-reading.csv", header=F, col.names=book.cols, 
                      stringsAsFactors=F, na.strings="")
    head(books, 1)
    dim(books)
```

### Genres

For each book title collected I also collected the names of the "shelves" it was added to. Many users sort their shelves by genre so this seems to be an adequate way to assign a single common genre to each title with a little processing.

First I'll find the most common shelf names and decide upon a set of genres to sort all the titles into, if possible.

```{r count_genres, cache=TRUE}
    require(stringr)
    genre.counts <- list()
    for (genre.set in books$genres) {
        genres <- str_extract_all(genre.set, "[a-z/-]+")[[1]]
        for (genre in genres) {
            if (!genre %in% names(genre.counts)) {
                genre.counts[genre] <- 1
            } else {
                genre.counts[genre] <- genre.counts[[genre]] + 1
            }
        }
    }
```
```{r examine_counts}
    sorted.counts <- sort(unlist(genre.counts),decreasing = T)
```
```{r}
    sorted.counts[1:100]
```
Now to pick a set of genres that most books will be able to be binned into.

```{r create_bins, cache=TRUE}
    genre.bins <- list("history"=c("history","american history","world history","european history","military history","world war i","wwi","wwii","world war ii","chinese history"),
                       "classics"=c("classics","classic"),
                       "historical fiction"=c("historical fiction"),
                       "fantasy"=c("fantasy"),
                       "biography"=c("biography","bio","biographies"),
                       "memoir"=c("memoir","autobiography","memoirs","biography memoir","biographies memoirs"),
                       "philosophy"=c("philosophy"),
                       "math and science"=c("science","psychology","sociology","anthropology","economics","mathematics","ecology","natural history","math","biology","statistics","social science","ethnography","library science","social sciences","physics","cognitive science","science nature","maths","astronomy"),
                       "mystery"=c("mystery","mysteries"),
                       "politics"=c("politics","political"),
                       "science fiction"=c("science fiction","sf","scifi","sci fi"),
                       "romance"=c("romance","romances"),
                       "young adult"=c("young adult","ya"),
                       "humor"=c("humor","comedy","humour"),
                       "religion"=c("religion","christianity","spirituality","religions","theology","islam","buddhism","biblical studies","bible studies","faith","christian living","church","catholic","religion spirituality","islamic books","christian non fiction","judaism","ministry","prayer","islamic","religious studies","buddhist"),
                       "business"=c("business","management","marketing","business books"),
                       "self improvement"=c("self help","self improvement","professional development","parenting","parenting books","health","fitness","running","exericse","self development","health fitness","personal development","motivation","health and fitness"),
                       "reference"=c("reference","art reference","writing reference"),
                       "thriller"=c("thriller","thrillers"),
                       "poetry"=c("poetry","poet","poetics"),
                       "horror"=c("horror"),
                       "adventure"=c("adventure"),
                       "literary fiction"=c("literary fiction","literary","lit fic","lit fiction"),
                       "food"=c("nutrition","foodie","cooking","food","cookbook","cookbooks","recipes","cook books","diets","wine"),
                       "childrens"=c("childrens","children","kid","kids","children s books"),
                       "technology"=c("technology","tech","programming","computer","computers","technical","web design","videogames","game design","gaming","science technology"),
                       "comics"=c("comics","comic","graphic novels","graphic novel"),
                       "the arts"=c("art","contemporary art","music related","art related","writing","music","design","film","architecture","writing books","literary criticism","photography","art books","cinema","theatre","art history","crafts","graphic design","drama","books on writing","plays","craft","art design","lit crit","theater","screenwriting","filmmaking","fashion","crafty","acting","literary theory","crafting","drawing","art and design","sewing","books on books","writing craft","musicology","film studies","jazz","music books","art photography","arts and crafts","design books","criticism and theory","composition","architecture and theory","opera","dance books"))
```
```{r get_genre, cache=TRUE}
    require(stringr)
    GetGenre <- function(shelf.set) {
        shelves <- str_extract_all(shelf.set, "[a-z/-]+")[[1]]
        for (s in shelves) {
            shelf <- str_trim(gsub("-"," ",tolower(s)))
            g <- names(grep(shelf,genre.bins,value=T))
            if (length(g)>0) return(g[1])
        }
        return(NA)
    }
```
```{r bin_books_by_genre, dependson="create_bins", cache=TRUE}
    books$genre <- sapply(books[,"genres"], GetGenre)
```
```{r examine_genre, cache=TRUE}
    table(books$genre, useNA = "ifany")
    genre.nas <- subset(books,is.na(genre))
    head(genre.nas$genres)
```
```{r count_genres_leftover, cache=TRUE, dependson="bin_books_by_genre"}
    require(stringr)
    genre.leftovers <- list()
    for (shelf.set in genre.nas$genres) {
        shelves <- str_extract_all(shelf.set, "[a-z/-]+")[[1]]
        for (shelf in shelves) {
            if (!shelf %in% names(genre.leftovers)) {
                genre.leftovers[shelf] <- 1
            } else {
                genre.leftovers[shelf] <- genre.leftovers[[shelf]] + 1
            }
        }
    }
```
Out of the 130396 books only 3752 were unable to be binned. The numbers look pretty reasonable, although 35902 historical fiction did surprise. 

I'll create a new CSV so I don't have to re-run this.

```{r write_books_with_genre, cache=TRUE}
    write.csv(books, file="data/books_with_genre.csv",row.names=F)
```

Obviously most of the interesting analysis is going to come from demographics, so now I'll merge in the user information

```{r merge_sets}
    merged <- merge(books, users)
    dim(merged)
```
That all looks like it should, so I'll write another csv

```{r write_merged, cache=TRUE}
    write.csv(merged, file="data/books_with_users.csv", row.names=F)
```

## Analysis

*Genres added to currently reading shelf over time*

When thinking about genre popularity and the book social network I think the most important date (out of date added to shelf, date started, and date finished) is the date the book was added to the shelf. This is the time when the user is interested in the book.

I'll also include gender, age, and location in this data set. I think these will provide interesting interactive features for the Shiny app.

```{r genre_time, cache=TRUE}
    books.users <- read.csv("data/books_with_users.csv")
    gt <- subset(books.users, !is.na(genre))
    dateadded <- as.Date(gt$dateadded, format="%a %b %d %H:%M:%S %z %Y")
    genre.time <- data.frame(dateadded, genre=gt$genre, gender=gt$gender, age=gt$age, location=gt$location)
    dim(genre.time)
```
```{r explore_genre_time}
    head(genre.time)
```
```{r write_genre_time, cache=TRUE}
    write.csv(genre.time, file="data/processed_books.csv", row.names=F)
```

## Geocoding

And of course we need to add latitude and longitude for each location.

```{r to_lat_lng}
    require(ggmap)
    ToLatLng <- function(loc) {
        geo <- geocode(loc)
        data.frame(loc=loc, lon=geo[1], lat=geo[2])
    }
```

```{r}
    genre.data <- read.csv("data/tmp_gd_ll.csv", stringsAsFactors=F)
    lls <- read.csv("data/tmp_lls.csv", stringsAsFactors=F)
```
```{r}
    for (i in 1:nrow(genre.data)) {
        location <- genre.data[i,"location"]
        if (is.na(genre.data[i,"lon"]) & location!="") {
            if (!location %in% lls$loc & !location %in% tried) {
                tried <- c(tried, location)
                ll <- ToLatLng(location)
                if (!is.na(ll$lon)) {
                    lls <- rbind(lls, ll)   
                }
            }
            l <- subset(lls,loc==location)
            if (nrow(l)==1) {
                genre.data[i,"lon"] <- l$lon
                genre.data[i,"lat"] <- l$lat
            }
        }
    }
```

```{r}
    write.csv(genre.data,"data/tmp_gd_ll.csv",row.names=F)
    write.csv(lls,"data/tmp_lls.csv",row.names=F)
    write.csv(tried, "data/tmp_tried.csv",row.names=F)
```
```{r}
    tried <- lls$loc
```


## Example Plots

Here are some quick plots to see some examples of what we'll be able to do with this data. These are quick and dirty, we'll save the fancy stuff for the shiny app.

```{r load_packages}
    require(ggplot2)
    require(dplyr)
    require(lubridate)
```

```{r data_for_plots, cache=TRUE}
   genre.data <- read.csv("data/processed_books.csv") 
```
One genre, historical fiction, by date

```{r hist_fic_by_date}
    hist.fic <- subset(genre.data, genre=="historical fiction")
    hist.fic$dateadded <- floor_date(as.Date(hist.fic$dateadded), "month")
    by_date <- hist.fic %.%
                group_by(dateadded) %.%
                summarise(counts=n())
    print(ggplot(by_date, aes(x=dateadded, y=counts)) + geom_line() + xlab("Date Added") + ylab("Number of Historical Fiction Titles Added"))
```
All genres by date
```{r all_by_date}
    genre.data$dateadded <- floor_date(as.Date(genre.data$dateadded), "month")
    by_genre_and_date <- genre.data %.%
                            group_by(genre, dateadded) %.%
                            summarise(counts=n())
    ggplot(by_genre_and_date, aes(x=dateadded, y=counts, color=genre, group=genre)) + geom_line()
```
Science Fiction by date and gender
```{r sf_date_gender}
    sf <- subset(genre.data, genre=="science fiction")
    sf_date_gender <- genre.data %.%
                        group_by(gender, dateadded) %.%
                        summarise(counts=n())
    ggplot(sf_date_gender, aes(x=dateadded, y=counts, color=gender, group=gender)) + geom_line()
```
Historical Fiction by location
```{r hf_map}
    has.ll <- subset(genre.data, !is.na(lon)&location!="")
```


## Rdata

we'll store the data frame in an Rdata file for easy loading in the shiny app

```{r}
    genre.data <- read.csv("data/processed_books.csv") 
    genre.data$dateadded <- as.Date(genre.data$dateadded)
    genre.data$monthadded <- floor_date(as.Date(genre.data$dateadded), "month")
    save(genre.data, file="data/genre_data.Rdata")
```


## Interactive Shiny App

Now to make it shiny! Visit the code in the ui.R and server.R files in the github repo and visit the shiny app at https://shanfu.shinyapps.io/shiny_books/







