---
title: "Books"
output: pdf_document
author: Shannon Rush
---

The goal of this analysis is to explore book genre popularity. Some questions I'll try to answer are: Have genre tastes changed over time? Does genre popularity have a pattern in the course of a year? Do demographics such as age or gender tend to report reading the same types of books?

# The Data

This data was collected via a public API from a popular book social network. All information is self reported, including demographic information, books read, and genre categorization. Unfortunately it is not necessary when reporting having read a book to categorize it by genre but fortunately many users do.

Two types of data were collected: User data, and information about books users self-reported to be "currently reading" 

A note about github: The collected CSV files are too large to share on github. However, you can see exactly how I collected the data at my repo called ['goodreads-analyses'](https://github.com/shannonrush/goodreads-analyses) I'm also happy to share the data I collected if you would like to reproduce my analyses or work with the data yourself. A word of caution: I collected this data too quickly for their tastes and ended up getting IP banned for about a month. Consume at your own risk. 

## User Data

```{r user_data, cache=TRUE}
    user.cols <- c("userID", "name", "gender", "age", "location", "lastactive", 
                   "readcount", "currentcount", "wantscount")
    users <- read.csv("data/users.csv", header=F, col.names=user.cols, stringsAsFactors=F)
    head(users)
    dim(users)
```

## Book Data

```{r book_data, cache=TRUE}
    book.cols <- c("userID", "bookID", "shelf", "dateadded", "datestarted", "datefinished", "title",
                   "isbn", "isbn13", "imageurl", "pages", "publisher", "publicationdate", "genres")
    books <- read.csv("data/currently-reading.csv", header=F, col.names=book.cols, 
                      stringsAsFactors=F, na.strings="")
    head(books, 1)
    dim(books)
```

### Genres

For each book title collected I also collected the names of the "shelves" it was added to. Many users sort their shelves by genre so this seems to be an adequate way to assign a single common genre to each title with a little processing.

First I'll find the most common shelf names and decide upon a set of genres to sort all the titles into, if possible.

```{r count_genres, cache=TRUE}
    require(stringr)
    genre.counts <- list()
    for (genre.set in books$genres) {
        genres <- str_extract_all(genre.set, "[a-z/-]+")[[1]]
        for (genre in genres) {
            if (!genre %in% names(genre.counts)) {
                genre.counts[genre] <- 1
            } else {
                genre.counts[genre] <- genre.counts[[genre]] + 1
            }
        }
    }
```
```{r examine_counts}
    sorted.counts <- sort(unlist(genre.counts),decreasing = T)
```
```{r}
    sorted.counts[1:100]
```
Now to pick a set of genres that most books will be able to be binned into.

```{r create_bins, cache=TRUE}
    genre.bins <- list("history"=c("history","american history","world history","european history","military history","world war i","wwi","wwii","world war ii","chinese history"),
                       "classics"=c("classics","classic"),
                       "historical fiction"=c("historical fiction"),
                       "fantasy"=c("fantasy"),
                       "biography"=c("biography","bio","biographies"),
                       "memoir"=c("memoir","autobiography","memoirs","biography memoir","biographies memoirs"),
                       "philosophy"=c("philosophy"),
                       "math and science"=c("science","psychology","sociology","anthropology","economics","mathematics","ecology","natural history","math","biology","statistics","social science","ethnography","library science","social sciences","physics","cognitive science","science nature","maths","astronomy"),
                       "mystery"=c("mystery","mysteries"),
                       "politics"=c("politics","political"),
                       "science fiction"=c("science fiction","sf","scifi","sci fi"),
                       "romance"=c("romance","romances"),
                       "young adult"=c("young adult","ya"),
                       "humor"=c("humor","comedy","humour"),
                       "religion"=c("religion","christianity","spirituality","religions","theology","islam","buddhism","biblical studies","bible studies","faith","christian living","church","catholic","religion spirituality","islamic books","christian non fiction","judaism","ministry","prayer","islamic","religious studies","buddhist"),
                       "business"=c("business","management","marketing","business books"),
                       "self improvement"=c("self help","self improvement","professional development","parenting","parenting books","health","fitness","running","exericse","self development","health fitness","personal development","motivation","health and fitness"),
                       "reference"=c("reference","art reference","writing reference"),
                       "thriller"=c("thriller","thrillers"),
                       "poetry"=c("poetry","poet","poetics"),
                       "horror"=c("horror"),
                       "adventure"=c("adventure"),
                       "literary fiction"=c("literary fiction","literary","lit fic","lit fiction"),
                       "food"=c("nutrition","foodie","cooking","food","cookbook","cookbooks","recipes","cook books","diets","wine"),
                       "childrens"=c("childrens","children","kid","kids","children s books"),
                       "technology"=c("technology","tech","programming","computer","computers","technical","web design","videogames","game design","gaming","science technology"),
                       "comics"=c("comics","comic","graphic novels","graphic novel"),
                       "the arts"=c("art","contemporary art","music related","art related","writing","music","design","film","architecture","writing books","literary criticism","photography","art books","cinema","theatre","art history","crafts","graphic design","drama","books on writing","plays","craft","art design","lit crit","theater","screenwriting","filmmaking","fashion","crafty","acting","literary theory","crafting","drawing","art and design","sewing","books on books","writing craft","musicology","film studies","jazz","music books","art photography","arts and crafts","design books","criticism and theory","composition","architecture and theory","opera","dance books"))
```
```{r get_genre, cache=TRUE}
    require(stringr)
    GetGenre <- function(shelf.set) {
        shelves <- str_extract_all(shelf.set, "[a-z/-]+")[[1]]
        for (s in shelves) {
            shelf <- str_trim(gsub("-"," ",tolower(s)))
            g <- names(grep(shelf,genre.bins,value=T))
            if (length(g)>0) return(g[1])
        }
        return(NA)
    }
```
```{r bin_books_by_genre, dependson="create_bins", cache=TRUE}
    books$genre <- sapply(books[,"genres"], GetGenre)
```
```{r examine_genre}
    table(books$genre, useNA = "ifany")
    genre.nas <- subset(books,is.na(genre))
    head(genre.nas$genres)
```
```{r count_genres_leftover, cache=TRUE, dependson="bin_books_by_genre"}
    require(stringr)
    genre.leftovers <- list()
    for (shelf.set in genre.nas$genres) {
        shelves <- str_extract_all(shelf.set, "[a-z/-]+")[[1]]
        for (shelf in shelves) {
            if (!shelf %in% names(genre.leftovers)) {
                genre.leftovers[shelf] <- 1
            } else {
                genre.leftovers[shelf] <- genre.leftovers[[shelf]] + 1
            }
        }
    }
```
Out of the 130396 books only 3752 were unable to be binned. The numbers look pretty reasonable, although 35902 historical fiction did surprise. 

I'll create a new CSV so I don't have to re-run this.

```{r write_books_with_genre, cache=TRUE}
    write.csv(books, file="data/books_with_genre.csv",row.names=F)
```

Obviously most of the interesting analysis is going to come from demographics, so now I'll merge in the user information

```{r merge_sets}
    merged <- merge(books, users)
    dim(merged)
```
That all looks like it should, so I'll write another csv

```{r write_merged, cache=TRUE}
    write.csv(merged, file="data/books_with_users.csv", row.names=F)
```





